MG7: Configurable and scalable 16S data analysis

### 2016-07-25 docs/mg7-paper/out/paper.tex
add a description of the first two steps in ref db generation

The starting point for our 16S reference database is RNA Central
\citep{rnacentral2014rnacentral},
\href{http://blog.rnacentral.org/2016/03/rnacentral-release-5.html}{version
5}.

\begin{quote}
Our 16S-DB7 Reference Database is a curated subset of sequences from the
NCBI nucleotide database \textbf{nt}. The sequences included were
selected by similarity with the bacterial and archaeal reference
sequences downloaded from the \textbf{RDP database}
\citep{cole2013ribosomal}. RDP unaligned sequences were used to capture
new 16S RNA sequences from \textbf{nt} using BLAST similarity search
strategies and then, performing additional curation steps to remove
sequences with poor taxonomic assignments to taxonomic nodes close to
the root of the taxonomy tree.

All the nucleotide sequences included in \textbf{nt} database has a
taxonomic assignment provided by the \textbf{Genbank} sequence
submitter. NCBI provides a table (available at
ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/) to do the mapping of any
Genbank Identifier (GI) to its Taxonomy Identifier (TaxID). Thus, we are
based on a crowdsourced submitter-maintained taxonomic annotation system
for reference sequences. It supposes a sustainable system able to face
the expected number of reference sequences that will populate the public
global nucleotide databases in the near future. Another advantageous
point is that we are based on NCBI taxonomy, the \emph{de-facto}
standard taxonomic classification for biomolecular data
\citep{cochrane20102010}. NCBI taxonomy is, undoubtedly, the most used
taxonomy all over the world and the most similar to the official
taxonomies of each specific field. This is a crucial point because all
the type-culture and tissue databanks follow this official taxonomical
classification and, in addition, all the knowledge accumulated during
last decades is referred to this taxonomy. In addition NCBI provides a
direct connection between taxonomical formal names and the physical
specimens that serve as exemplars for the species
\citep{federhen2014type}.

Certainly, if metagenomics results are easily integrated with the
theoretical and experimental knowledge of each specific area, the impact
of metagenomics will be higher than if it progresses as a disconnected
research branch. Considering that metagenomics data interoperability,
which is especially critical in clinical environments, requires a stable
taxonomy to be used as reference, we decided to rely on the most widely
used taxonomy: the NCBI taxonomy. In addition, the biggest global
sequence database GenBank follows this taxonomy to register the origin
of all their submitted sequences. Our 16S database building strategy
allows the substitution of the 16S database by any other subset of
\textbf{nt}, even by the complete \textbf{nt} database if it would be
needed, for example, for analyzing shotgun metagenomics data. This
possibility of changing the reference database provides flexibility to
the system enabling it for easy updating and project-driven
personalization.
\end{quote}

>>>>

The starting point for our 16S reference database is RNA Central
\citep{rnacentral2014rnacentral},
\href{http://blog.rnacentral.org/2016/03/rnacentral-release-5.html}{version
5}. RNACentral was chosen being the most up to date, comprehensive RNA
sequence repository, including among others all RNA data from Silva,
GreenGenes, RDP, ENA, and RefSeq. First we take those sequences which

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  are annotated as being of \texttt{rRNA} type\footnote{We are aware of
    the existence of a gene annotation corresponding to 16S, that we are
    \textbf{not using} due to a significant amount of 16S sequences
    lacking the corresponding annotation}
\item
  their length is at least \texttt{1300}\footnote{Note that 16S
    sequences are sometimes part of an entry corresponding to whole
    small subunits; that's why we do not set a maximum length threshold}
\item
  have at least one taxonomic assignment to a descendant of
  \emph{Bacteria} or \emph{Archaea}
\item
  their lineage does not contain a set of taxa deemed
  uninformative\footnote{for example: ``unclassified Bacteria
    (miscellaneous)'' or ``unclassified''. Assigning a read to an
    ``unclassified'' taxon defeats the first and foremost goal of
    taxonomic profiling: \emph{classification}.}
\end{enumerate}

After this first step, we drop redundant assignments: if sequences
\(S_1 \subseteq S_2\) share an assignment \(T\), it gets dropped from
\(S_1\); sequences which as a result end up having no assignments are
removed.

Once we have non-redundant set of RNA sequences containing all those
corresponding to 16S, we can apply our procedure for checking internal
consistency of taxonomic assignments.
